---
title: "Neutrophil Transcriptomic Signature in COVID-19"
subtitle: "Differential Expression Analysis of GSE212041 | COVID-19 Day 0 vs Healthy Controls"
author: "Bastien FRANJA"
date: "`r format(Sys.time(), '%B %d, %Y')`"
institute: "EFREI - Genomics, Epigenetics and Applications ST2GEA-2526PSP01"
instructor: "Professor Mano Mathew"
student_id: "20221198"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
    fig_width: 10
    fig_height: 7
---


```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo    = TRUE,       # Show all code
  message = FALSE,      # Suppress package messages
  warning = FALSE,      # Suppress warnings
  cache   = FALSE,      # No caching (ensures full reproducibility)
  dpi     = 150,
  fig.align = "center"
)
```

---

# Introduction

## Biological Background

Neutrophils are the most abundant circulating leukocytes and serve as a first line of defense against infection. In COVID-19 (caused by SARS-CoV-2), neutrophils play a dual and paradoxical role:

- **Protective**: Early recruitment and pathogen clearance via phagocytosis, degranulation, and NETosis (Neutrophil Extracellular Traps)
- **Pathological**: Excessive activation drives cytokine storm, acute lung injury (ALI), and thrombo-inflammation

Understanding the transcriptomic program of neutrophils at hospital admission (Day 0) may help identify **early biomarkers of disease severity** and **therapeutic targets** to modulate neutrophil hyper-activation.

## Dataset

**Source**: GEO accession [GSE212041](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE212041)

**Publication**: *"Longitudinal characterization of circulating neutrophils uncovers phenotypes associated with severity in hospitalized COVID-19 patients"* (2022)

| Feature | Value |
|---|---|
| Subjects | 306 COVID-19 patients + 86 controls |
| Samples | ~781 bulk RNA-seq samples |
| Time points | D0 (admission), D3, D7, DE (event-driven) |
| Platform | Illumina NextSeq 500 |
| Genes | 60,640 in raw matrix |

## Biological Question

> **"What neutrophil transcriptomic signature distinguishes hospitalized COVID-19 patients from healthy controls at the time of hospital admission?"**

**Comparison**: COVID-19 Day 0 patients (`covid_day0`) vs Healthy Controls (`healthy_control`)

---

# Environment Setup

```{r install-packages, eval=FALSE}
# Run this block ONCE to install all required packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

pkgs <- c(
  "DESeq2", "org.Hs.eg.db", "AnnotationDbi", "clusterProfiler",
  "enrichplot", "ggplot2", "ggrepel", "pheatmap",
  "RColorBrewer", "dplyr", "tidyr", "tibble", "scales"
)
for (pkg in pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE))
    BiocManager::install(pkg, ask = FALSE)
}
```

```{r load-libraries}
# Load all required libraries
suppressPackageStartupMessages({
  library(DESeq2)          # Core: differential expression (negative binomial model)
  library(org.Hs.eg.db)    # Human gene annotation database
  library(AnnotationDbi)   # Interface to annotation databases
  library(clusterProfiler) # GO and KEGG enrichment analysis
  library(enrichplot)      # Visualization for enrichment results
  library(ggplot2)         # Publication-quality plots
  library(ggrepel)         # Non-overlapping text labels
  library(pheatmap)        # Clustered heatmaps
  library(RColorBrewer)    # Color palettes
  library(dplyr)           # Data manipulation
  library(tidyr)           # Data reshaping
  library(tibble)          # Modern data frames
  library(scales)          # Axis formatting
})

# Mask AnnotationDbi::select to avoid namespace conflict with dplyr
select  <- dplyr::select
filter  <- dplyr::filter
rename  <- dplyr::rename
mutate  <- dplyr::mutate
arrange <- dplyr::arrange

# Output folders
dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)
dir.create("results/tables",  recursive = TRUE, showWarnings = FALSE)

cat("All libraries loaded successfully.\n")
```

---

# Data Import and Preprocessing

## Load Count Matrix

```{r load-data}
count_data <- read.table(
  "/Users/monmacbook/Downloads/data/GSE212041_Neutrophil_RNAseq_Count_Matrix.txt.gz",
  header      = TRUE,
  row.names   = 1,
  check.names = FALSE,
  sep         = "\t"
)

# Remove Symbol column if present
if (colnames(count_data)[1] %in% c("Symbol", "Gene") ||
    grepl("Gene", colnames(count_data)[1])) {
  count_data <- count_data[, -1]
  cat("Removed metadata column from matrix.\n")
}

cat("Count matrix dimensions:\n")
cat("  Genes  :", nrow(count_data), "\n")
cat("  Samples:", ncol(count_data), "\n")

# Affiche les 3 premières colonnes pour vérifier la structure
cat("Premières colonnes:", paste(colnames(count_data)[1:5], collapse = ", "), "\n")

# Quick sanity check: counts should be non-negative integers
cat("\nFirst 3 rows and columns of raw count matrix:\n")
print(count_data[1:3, 1:3])
```

## Build Sample Metadata

```{r build-metadata}
# The sample naming convention encodes timepoint and condition:
#   - "[ID]_D0"  = COVID-19 patient at hospital admission (Day 0) ← our focus
#   - "[ID]_D3"  = Day 3 post-admission
#   - "[ID]_D7"  = Day 7 post-admission
#   - "[ID]_DE"  = Event-driven draw (intubation/extubation)
#   - "H[N]_H"   = Healthy control
#
# We derive condition labels directly from sample names (no external metadata file needed)

sample_names <- colnames(count_data)

metadata <- data.frame(
  sample_id = sample_names,
  condition = "covid_other",   # default for unclassified timepoints
  stringsAsFactors = FALSE
)

for (i in seq_along(sample_names)) {
  name <- sample_names[i]
  if      (grepl("_H$|^H[0-9]+_H", name)) metadata$condition[i] <- "healthy_control"
  else if (grepl("_D0$",           name)) metadata$condition[i] <- "covid_day0"
  else if (grepl("_D3$",           name)) metadata$condition[i] <- "covid_day3"
  else if (grepl("_D7$",           name)) metadata$condition[i] <- "covid_day7"
  else if (grepl("_DE$",           name)) metadata$condition[i] <- "covid_event"
}

cat("Full sample distribution across all conditions:\n")
print(table(metadata$condition))
```

## Filter to Comparison Groups

```{r filter-groups}
# We restrict analysis to:
#   - COVID-19 Day 0  : Earliest available time point after hospitalization
#   - Healthy control : Infection-naive baseline reference
#
# This addresses the key biological question: what is unique about
# COVID-19 neutrophils at the time of admission?

metadata_clean <- metadata[
  metadata$condition %in% c("healthy_control", "covid_day0"), 
]

# Subset the count matrix to the selected samples
count_data_clean <- count_data[, metadata_clean$sample_id, drop = FALSE]

# Set reference level: log2FC > 0 means HIGHER in COVID-19 relative to healthy
metadata_clean$condition <- factor(
  metadata_clean$condition,
  levels = c("healthy_control", "covid_day0")
)

cat("Analysis comparison:\n")
cat("  Healthy controls  :", sum(metadata_clean$condition == "healthy_control"), "samples\n")
cat("  COVID-19 Day 0    :", sum(metadata_clean$condition == "covid_day0"), "samples\n")
cat("\n  Reference level (denominator): healthy_control\n")
cat("  Test level (numerator)        : covid_day0\n")
cat("  → Positive log2FC = higher in COVID-19\n")
```

---

# Differential Expression Analysis (DESeq2)

## Create DESeq2 Object

```{r create-dds}
# DESeqDataSetFromMatrix requires:
#   - countData : integer matrix of raw counts (no normalization!)
#   - colData   : sample metadata aligned with count matrix columns
#   - design    : formula specifying the experimental design

# round() ensures counts are integers (required by DESeq2)
dds <- DESeqDataSetFromMatrix(
  countData = round(count_data_clean),
  colData   = metadata_clean,
  design    = ~ condition    # Simple two-group comparison
)

# Pre-filtering: remove genes with fewer than 10 counts in fewer than 3 samples
# Rationale: Low-count genes are uninformative and inflate multiple testing burden
keep <- rowSums(counts(dds) >= 10) >= 3
dds  <- dds[keep, ]

cat("After low-count filtering:\n")
cat("  Genes retained:", nrow(dds), "\n")
cat("  Genes removed :", sum(!keep), "\n")
```

## Run DESeq2 Pipeline

```{r run-deseq2}
# DESeq() performs three sequential steps:
#   1. estimateSizeFactors()  — library size normalization
#   2. estimateDispersions()  — gene-wise dispersion estimation
#   3. nbinomWaldTest()       — Wald test for each gene

dds <- DESeq(dds)

# Extract results for the covid_day0 vs healthy_control contrast
# alpha = 0.05 sets the independent filtering threshold
res <- results(
  dds,
  contrast      = c("condition", "covid_day0", "healthy_control"),
  alpha         = 0.05,
  pAdjustMethod = "BH"   # Benjamini-Hochberg multiple testing correction
)

cat("DESeq2 analysis summary:\n")
summary(res)
```

## Annotate Results

```{r annotate}
# Convert results to a data frame and filter out genes with NA padj
# (NA occurs when DESeq2's independent filtering excludes very low-count genes)
res_df <- as.data.frame(res) %>%
  tibble::rownames_to_column("ensembl_id") %>%
  dplyr::filter(!is.na(padj), !is.na(log2FoldChange))

# Map Ensembl IDs → Gene Symbol + Entrez ID using org.Hs.eg.db
gene_anno <- suppressMessages(
  AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = res_df$ensembl_id,
    columns = c("ENSEMBL", "SYMBOL", "ENTREZID"),
    keytype = "ENSEMBL"
  )
) %>%
  dplyr::distinct(ENSEMBL, .keep_all = TRUE) %>%
  dplyr::rename(ensembl_id = ENSEMBL)

# Merge annotation with DE results and classify each gene
res_ann <- res_df %>%
  dplyr::left_join(gene_anno, by = "ensembl_id") %>%
  dplyr::mutate(
    # Fall back to Ensembl ID when symbol is unavailable
    symbol = dplyr::if_else(!is.na(SYMBOL), SYMBOL, ensembl_id),

    # Classification thresholds: |log2FC| > 1 (2-fold) and FDR < 5%
    direction = dplyr::case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "UP",
      padj < 0.05 & log2FoldChange < -1 ~ "DOWN",
      TRUE                               ~ "NS"
    ),

    # For volcano plot y-axis
    neg_log10_padj = -log10(padj)
  )

# Report DEG counts
n_up   <- sum(res_ann$direction == "UP")
n_down <- sum(res_ann$direction == "DOWN")
n_ns   <- sum(res_ann$direction == "NS")

cat("DEG classification (|log2FC| > 1, FDR < 0.05):\n")
cat("  Upregulated  :", n_up,   "genes\n")
cat("  Downregulated:", n_down, "genes\n")
cat("  Non-significant:", n_ns, "genes\n")

# Save full annotated results table
write.csv(res_ann, "results/tables/DEG_annotated_results.csv", row.names = FALSE)
cat("\nFull results saved to: results/tables/DEG_annotated_results.csv\n")
```

```{r top-genes-table}
# Display top 15 DEGs by adjusted p-value for quick inspection
cat("Top 15 most significant DEGs:\n")

res_ann %>%
  dplyr::filter(direction != "NS") %>%
  dplyr::arrange(padj) %>%
  dplyr::select(symbol, log2FoldChange, baseMean, padj, direction) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::mutate(
    log2FoldChange = round(log2FoldChange, 2),
    baseMean       = round(baseMean, 1),
    padj           = formatC(padj, format = "e", digits = 2)
  ) %>%
  knitr::kable(
    caption = "Top 15 differentially expressed genes (sorted by FDR)",
    col.names = c("Gene", "log2FC", "Mean Count", "Adj. P-value", "Direction")
  )
```

---

# Quality Control and Visualisation

## Variance Stabilising Transformation (VST)

```{r vst}
# VST transforms raw counts so that variance is approximately constant 
# across expression levels. This is required for PCA and heatmaps because:
#   - Highly expressed genes would dominate without transformation
#   - Log2(counts) alone is unstable at low counts
#   - blind = FALSE uses the experimental design to improve dispersion estimates

vst_data <- vst(dds, blind = FALSE)
cat("VST transformation complete.\n")
cat("  Dimensions:", nrow(assay(vst_data)), "genes x", ncol(assay(vst_data)), "samples\n")
```

## PCA — Sample Clustering

```{r pca, fig.cap="PCA of VST-normalized counts. Each point is one sample; ellipses show 95% confidence regions per group."}
# PCA reduces the high-dimensional gene expression matrix to 2 axes.
# Well-separated groups → consistent transcriptional differences.
# Within-group spread → biological variability.

pca_data  <- plotPCA(vst_data, intgroup = "condition", returnData = TRUE)
pct_var   <- round(100 * attr(pca_data, "percentVar"), 1)

ggplot(pca_data, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(fill = condition), alpha = 0.1, geom = "polygon") +
  scale_color_manual(
    values = c("covid_day0" = "#E41A1C", "healthy_control" = "#377EB8"),
    labels = c("COVID-19 Day 0", "Healthy Control")
  ) +
  scale_fill_manual(
    values = c("covid_day0" = "#E41A1C", "healthy_control" = "#377EB8"),
    guide  = "none"
  ) +
  labs(
    title    = "PCA — Sample Clustering by Condition",
    subtitle = paste0("Top 500 most variable genes | PC1: ", pct_var[1],
                      "% | PC2: ", pct_var[2], "% variance explained"),
    x     = paste0("PC1 (", pct_var[1], "%)"),
    y     = paste0("PC2 (", pct_var[2], "%)"),
    color = "Group"
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40")
  )
```

**Interpretation**: PC1 (the axis capturing the most variance) should clearly separate COVID-19 patients from healthy controls, confirming that infection drives a strong, reproducible transcriptional shift in circulating neutrophils.

## Sample-to-Sample Distance Heatmap

```{r sample-distance, fig.cap="Euclidean distances between VST-normalized samples. Darker = more similar. Ward clustering."}
# Sample distance heatmap is a complementary QC to PCA.
# It shows pairwise similarity across ALL genes, not just the top 500.
# We expect healthy controls to cluster tightly together and
# COVID-19 samples to form a distinct cluster.

samp_dist <- dist(t(assay(vst_data)))
samp_mat  <- as.matrix(samp_dist)

# Build informative row/column names
sample_labels <- paste(
  dplyr::if_else(vst_data$condition == "covid_day0", "COVID", "Healthy"),
  seq_len(ncol(samp_mat)),
  sep = "_"
)
rownames(samp_mat) <- sample_labels
colnames(samp_mat) <- sample_labels

# Annotation sidebar
samp_anno <- data.frame(
  Group = dplyr::if_else(
    vst_data$condition == "covid_day0", "COVID-19 Day 0", "Healthy Control"
  ),
  row.names = sample_labels
)

anno_colors <- list(
  Group = c("COVID-19 Day 0" = "#E41A1C", "Healthy Control" = "#377EB8")
)

pheatmap(
  mat               = samp_mat,
  color             = colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Blues")))(100),
  annotation_col    = samp_anno,
  annotation_row    = samp_anno,
  annotation_colors = anno_colors,
  clustering_method = "ward.D2",
  show_rownames     = TRUE,
  show_colnames     = FALSE,
  fontsize          = 9,
  border_color      = "white",
  main              = "Sample-to-Sample Distance\n(VST-normalized | Ward clustering)"
)
```

---

# Visualisation of DE Results

## Volcano Plot

```{r volcano, fig.cap="Volcano plot. Red = upregulated in COVID-19, blue = downregulated. Dashed lines: |log2FC|=1 and FDR=0.05."}
# Select top 15 DEGs to label (by adjusted p-value)
top_ids <- res_ann %>%
  dplyr::filter(direction != "NS") %>%
  dplyr::arrange(padj) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::pull(ensembl_id)

res_ann <- res_ann %>%
  dplyr::mutate(plot_label = dplyr::if_else(ensembl_id %in% top_ids, symbol, ""))

# Color scheme consistent across all figures
deg_colors <- c("UP" = "#E41A1C", "DOWN" = "#377EB8", "NS" = "grey70")

ggplot(
  res_ann %>% dplyr::arrange(direction),
  aes(x = log2FoldChange, y = neg_log10_padj,
      color = direction, size = direction, alpha = direction)
) +
  geom_point(shape = 16, stroke = 0.2) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "black", linewidth = 0.5) +
  ggrepel::geom_text_repel(
    data          = dplyr::filter(res_ann, plot_label != ""),
    aes(label     = plot_label),
    color         = "black",
    size          = 3,
    fontface      = "bold",
    box.padding   = 0.4,
    point.padding = 0.3,
    max.overlaps  = 30,
    segment.color = "grey50",
    segment.size  = 0.3
  ) +
  scale_color_manual(
    values = deg_colors,
    labels = c(
      UP   = paste0("Upregulated (", n_up, ")"),
      DOWN = paste0("Downregulated (", n_down, ")"),
      NS   = paste0("Not significant (", n_ns, ")")
    )
  ) +
  scale_size_manual(values  = c(UP = 1.8, DOWN = 1.8, NS = 0.8)) +
  scale_alpha_manual(values = c(UP = 0.9, DOWN = 0.9, NS = 0.35)) +
  labs(
    title    = "Volcano Plot — COVID-19 Day 0 vs Healthy Controls",
    subtitle = paste0(
      n_up + n_down, " DEGs | |log2FC| > 1, FDR < 0.05 | Labeled: top 15 by FDR"
    ),
    x     = expression(log[2]~"Fold Change (COVID-19 / Healthy)"),
    y     = expression(-log[10]~"Adjusted P-value"),
    color = "Direction"
  ) +
  guides(size = "none", alpha = "none") +
  theme_bw(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(size = 10, color = "grey40"),
    panel.grid.minor = element_blank()
  )
```

**Gene selection justification**: Labels show the 15 genes with the lowest FDR-adjusted p-value among DEGs, ensuring the most statistically robust signals are highlighted regardless of fold-change magnitude.

## Heatmap — Top 50 DEGs

```{r heatmap, fig.height=14, fig.cap="Hierarchical clustering heatmap of the top 50 DEGs (ranked by FDR). Z-scored VST values. Ward.D2 clustering."}
# Select top 50 DEGs (most significant)
top50_ids <- res_ann %>%
  dplyr::filter(direction != "NS") %>%
  dplyr::arrange(padj) %>%
  dplyr::slice_head(n = 50) %>%
  dplyr::pull(ensembl_id)

# Fallback if very few DEGs found
if (length(top50_ids) < 10) {
  top50_ids <- res_ann %>%
    dplyr::arrange(padj) %>%
    dplyr::slice_head(n = 50) %>%
    dplyr::pull(ensembl_id)
  cat("Warning: fewer than 10 DEGs; using top 50 by p-value instead.\n")
}

# Extract VST matrix for selected genes
avail_ids <- top50_ids[top50_ids %in% rownames(assay(vst_data))]
vst_mat   <- assay(vst_data)[avail_ids, ]

# Replace Ensembl IDs with gene symbols for readability
sym_map   <- res_ann %>%
  dplyr::filter(ensembl_id %in% avail_ids) %>%
  dplyr::select(ensembl_id, symbol) %>%
  dplyr::distinct(ensembl_id, .keep_all = TRUE)

row_idx           <- match(rownames(vst_mat), sym_map$ensembl_id)
rownames(vst_mat) <- make.unique(
  dplyr::if_else(!is.na(row_idx), sym_map$symbol[row_idx], rownames(vst_mat)),
  sep = "_"
)

# Z-score normalization per gene (row-wise)
# Allows comparison of expression patterns regardless of absolute expression level
vst_scaled <- t(scale(t(vst_mat)))
vst_scaled[vst_scaled >  3] <-  3   # Cap extreme values for better color contrast
vst_scaled[vst_scaled < -3] <- -3

# Column annotation: group and condition label
col_anno <- data.frame(
  Group = dplyr::if_else(
    colnames(vst_scaled) %in%
      metadata_clean$sample_id[metadata_clean$condition == "covid_day0"],
    "COVID-19 Day 0", "Healthy Control"
  ),
  row.names = colnames(vst_scaled)
)

anno_colors_ht <- list(
  Group = c("COVID-19 Day 0" = "#E41A1C", "Healthy Control" = "#377EB8")
)

pheatmap(
  mat                      = vst_scaled,
  color                    = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  breaks                   = seq(-3, 3, length.out = 101),
  annotation_col           = col_anno,
  annotation_colors        = anno_colors_ht,
  cluster_rows             = TRUE,
  cluster_cols             = TRUE,
  clustering_method        = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  show_rownames            = TRUE,
  show_colnames            = FALSE,
  fontsize_row             = 8,
  fontsize                 = 10,
  border_color             = NA,
  main = "Top 50 DEGs | COVID-19 Day 0 vs Healthy Controls\nZ-scored VST Counts | Ward.D2 Clustering"
)
```

## Top 10 DEG Expression Boxplots

```{r boxplots, fig.height=12, fig.cap="Normalized expression of the 10 most significant DEGs. Jittered points show individual samples."}
# Extract top 10 DEGs
top10_ids <- res_ann %>%
  dplyr::filter(direction != "NS") %>%
  dplyr::arrange(padj) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::pull(ensembl_id)

# Build long-format data frame for ggplot
norm_counts <- counts(dds, normalized = TRUE)

box_df <- norm_counts[
  top10_ids[top10_ids %in% rownames(norm_counts)], ,
  drop = FALSE
] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensembl_id") %>%
  tidyr::pivot_longer(
    cols      = -ensembl_id,
    names_to  = "sample",
    values_to = "count"
  ) %>%
  dplyr::left_join(metadata_clean, by = c("sample" = "sample_id")) %>%
  dplyr::left_join(dplyr::select(res_ann, ensembl_id, symbol, direction),
                   by = "ensembl_id") %>%
  dplyr::mutate(
    log2_count = log2(count + 1),          # Log-transform to reduce skew
    group      = dplyr::if_else(
      condition == "covid_day0", "COVID-19", "Healthy"
    ),
    gene_label = paste0(symbol, " (", direction, ")")
  )

ggplot(box_df, aes(x = group, y = log2_count, fill = group, color = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, linewidth = 0.5) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6, stroke = 0.2) +
  facet_wrap(~ gene_label, scales = "free_y", ncol = 2) +
  scale_fill_manual(values  = c("COVID-19" = "#E41A1C", "Healthy" = "#377EB8")) +
  scale_color_manual(values = c("COVID-19" = "#C0392B", "Healthy" = "#2471A3")) +
  labs(
    title    = "Top 10 DEGs — Individual Gene Expression",
    subtitle = expression(log[2]~"(normalized count + 1)"),
    x        = NULL,
    y        = expression(log[2]~"(norm count + 1)")
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    strip.text       = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "grey92"),
    legend.position  = "none",
    axis.text.x      = element_text(size = 10)
  )
```

---

# Functional Enrichment Analysis

## Prepare Gene Lists

```{r enrichment-setup}
# Extract Entrez IDs for enrichment (clusterProfiler requires Entrez, not Ensembl)
up_entrez   <- res_ann %>%
  dplyr::filter(direction == "UP",   !is.na(ENTREZID)) %>%
  dplyr::pull(ENTREZID)

down_entrez <- res_ann %>%
  dplyr::filter(direction == "DOWN", !is.na(ENTREZID)) %>%
  dplyr::pull(ENTREZID)

universe    <- res_ann %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::pull(ENTREZID)

cat("Genes available for enrichment:\n")
cat("  Upregulated genes (with Entrez ID)  :", length(up_entrez), "\n")
cat("  Downregulated genes (with Entrez ID):", length(down_entrez), "\n")
cat("  Background universe                 :", length(universe), "\n")
```

## GO Biological Process — Upregulated Genes

```{r go-up, fig.cap="GO Biological Process enrichment in upregulated genes. Dot size = gene ratio; color = adjusted p-value."}
# Gene Ontology (GO) enrichment tests whether a set of genes is statistically
# over-represented in any functional category.
# 
# We test the "Biological Process" (BP) ontology, which describes the
# broader cellular and physiological roles that genes participate in.
# pvalueCutoff = 0.05 applies Benjamini-Hochberg correction internally.

if (length(up_entrez) >= 5) {
  go_up <- suppressMessages(
    clusterProfiler::enrichGO(
      gene          = up_entrez,
      universe      = universe,
      OrgDb         = org.Hs.eg.db,
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.20,
      readable      = TRUE   # converts Entrez IDs to gene symbols in output
    )
  )

  if (!is.null(go_up) && nrow(go_up@result) > 0) {
    enrichplot::dotplot(go_up, showCategory = 20) +
      labs(
        title    = "GO Biological Process — Upregulated Genes",
        subtitle = "COVID-19 Day 0 vs Healthy Controls | FDR < 0.05"
      ) +
      theme_bw(base_size = 11) +
      theme(
        plot.title  = element_text(face = "bold"),
        axis.text.y = element_text(size = 9)
      )
  } else {
    cat("No significant GO terms found for upregulated genes.\n")
  }
} else {
  cat("Too few upregulated genes for GO enrichment.\n")
}
```

## GO Biological Process — Downregulated Genes

```{r go-down, fig.cap="GO Biological Process enrichment in downregulated genes."}
if (length(down_entrez) >= 5) {
  go_down <- suppressMessages(
    clusterProfiler::enrichGO(
      gene          = down_entrez,
      universe      = universe,
      OrgDb         = org.Hs.eg.db,
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.20,
      readable      = TRUE
    )
  )

  if (!is.null(go_down) && nrow(go_down@result) > 0) {
    enrichplot::dotplot(go_down, showCategory = 20) +
      labs(
        title    = "GO Biological Process — Downregulated Genes",
        subtitle = "COVID-19 Day 0 vs Healthy Controls | FDR < 0.05"
      ) +
      theme_bw(base_size = 11) +
      theme(
        plot.title  = element_text(face = "bold"),
        axis.text.y = element_text(size = 9)
      )
  } else {
    cat("No significant GO terms found for downregulated genes.\n")
  }
} else {
  cat("Too few downregulated genes for GO enrichment.\n")
}
```

## KEGG Pathway Enrichment

```{r kegg-up, fig.cap="KEGG pathway enrichment in upregulated genes. Each dot = one pathway."}
# KEGG (Kyoto Encyclopedia of Genes and Genomes) tests enrichment in curated
# biological pathways (e.g., "Neutrophil extracellular trap formation",
# "Cytokine-cytokine receptor interaction", "JAK-STAT signaling").

if (length(up_entrez) >= 5) {
  kegg_up <- suppressMessages(
    clusterProfiler::enrichKEGG(
      gene          = up_entrez,
      universe      = universe,
      organism      = "hsa",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.20
    )
  )

  if (!is.null(kegg_up) && nrow(kegg_up@result) > 0) {
    enrichplot::dotplot(kegg_up, showCategory = 20) +
      labs(
        title    = "KEGG Pathways — Upregulated Genes",
        subtitle = "COVID-19 Day 0 vs Healthy Controls | FDR < 0.05"
      ) +
      theme_bw(base_size = 11) +
      theme(
        plot.title  = element_text(face = "bold"),
        axis.text.y = element_text(size = 9)
      )
  } else {
    cat("No significant KEGG pathways found for upregulated genes.\n")
  }
}
```

```{r kegg-down, fig.cap="KEGG pathway enrichment in downregulated genes."}
if (length(down_entrez) >= 5) {
  kegg_down <- suppressMessages(
    clusterProfiler::enrichKEGG(
      gene          = down_entrez,
      universe      = universe,
      organism      = "hsa",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.20
    )
  )

  if (!is.null(kegg_down) && nrow(kegg_down@result) > 0) {
    enrichplot::dotplot(kegg_down, showCategory = 20) +
      labs(
        title    = "KEGG Pathways — Downregulated Genes",
        subtitle = "COVID-19 Day 0 vs Healthy Controls | FDR < 0.05"
      ) +
      theme_bw(base_size = 11) +
      theme(
        plot.title  = element_text(face = "bold"),
        axis.text.y = element_text(size = 9)
      )
  } else {
    cat("No significant KEGG pathways found for downregulated genes.\n")
  }
}
```

---

# Discussion of Main Findings

## Key Transcriptional Signatures

The DESeq2 analysis identified **`r n_up + n_down` differentially expressed genes** (FDR < 0.05, |log2FC| > 1) between COVID-19 patients at admission and healthy controls. The results point to four major biological themes:

### 1. Neutrophil Hyperactivation and NETosis

- **CD177** (log2FC ≈ +6.4) is a highly specific marker of activated neutrophils that facilitates trans-endothelial migration. Its dramatic upregulation confirms a state of systemic neutrophil priming in COVID-19.
- **DEFA1/DEFA3** (log2FC > +20) encode alpha-defensins, which are key structural components of Neutrophil Extracellular Traps (NETs). NETosis has been implicated in COVID-19-associated thrombosis, endothelial damage, and cytokine amplification.

### 2. Interferon Antiviral Response

- **IFI27** (log2FC ≈ +5.3), an interferon-stimulated gene, reflects active type I IFN signaling. While essential for antiviral defense against SARS-CoV-2, excessive IFN-I responses in neutrophils can paradoxically worsen inflammation and tissue damage.
- The interferon signature may also explain some clinical observations in COVID-19, such as immune paralysis in severe disease.

### 3. Acute Phase and Innate Immune Activation

- **HP (Haptoglobin)** (log2FC ≈ +3.8) is an acute phase protein and indicator of systemic inflammation. Its elevation in neutrophils suggests transcriptional rewiring toward an emergency granulopoiesis state.
- **TLR5** upregulation indicates enhanced innate pattern recognition capacity, likely to amplify responses against microbial-associated molecular patterns (MAMPs).

### 4. Downregulation of Ribosomal Pseudogenes

Many downregulated genes appear to be **ribosomal protein pseudogenes**. This pattern likely reflects either:
- Transcriptional reprogramming during the emergency granulopoiesis response
- Cellular stress responses that suppress normal housekeeping gene expression
- Technical noise specific to the large imbalance in sample sizes (374 COVID vs. 8 healthy)

## Clinical Implications

| Finding | Potential Clinical Relevance |
|---|---|
| CD177 upregulation | Possible serum biomarker for neutrophil hyperactivation |
| DEFA1/3 overexpression | Suggests NET formation; supports anti-NET strategies (e.g., DNase, PAD4 inhibitors) |
| IFI27 / Interferon response | Supports use of IFN-modulating therapies in early COVID-19 |
| Acute phase signature (HP, SAA) | Consistent with clinical CRP/ferritin elevation in severe cases |

## Limitations

1. **Imbalanced group sizes**: 374 COVID-19 vs. only 8 healthy controls — this imbalance can inflate fold-change estimates and reduce statistical power for detecting downregulated genes.
2. **Lack of clinical stratification**: All COVID-19 Day 0 patients are pooled regardless of severity (ICU vs. non-ICU). Sub-group analysis could reveal severity-specific signatures.
3. **Bulk RNA-seq**: Cannot resolve neutrophil subpopulations (low-density vs. mature vs. immature). Single-cell approaches would be more informative.
4. **No batch correction**: The analysis does not account for potential batch effects across sequencing runs, which may inflate variance in this large dataset.

---

# Conclusion

> **COVID-19 neutrophils at hospital admission exhibit a transcriptional signature dominated by hyperactivation (CD177, DEFA1), interferon response (IFI27), and acute-phase inflammation (HP, TLR5) — consistent with a state of emergency granulopoiesis and NETosis-prone phenotype.**

This signature is biologically coherent with the known immunopathology of severe COVID-19 and identifies several actionable targets for therapeutic intervention.

---

# Session Information

```{r session-info}
# Ensure full reproducibility: document all package versions
sessionInfo()
```




